{ Das CrypTool-Skript: \\ Kryptographie, Mathematik und mehr} \\ \vspace{30pt} {\large Hintergrundmaterial und Zusatzinformationen\\%*[-8pt] zum freien E-Learning-Programm CrypTool\\ (9. Auflage -- zu Version 1.4.20)} \vspace{60pt}

### 2009-07-09 script/de/chapters/numbertheory.tex
chapter on number theory: add Sage examples for calculating primitive roots; fixed duplicate } introduced in r2699 resulting in missing chapter numbers; TODO: translate

\newpage
\hypertarget{Chapter_ElementaryNT}{}
\chapter{Einführung in die elementare Zahlentheorie mit Beispielen}
\label{Chapter_ElementaryNT}
(Bernhard Esslinger, Juli 2001; Updates: Nov. 2001, Juni 2002, Mai 2003,
 Mai 2005, März 2006, Juni 2007) \\

@@

{\bf Beispiel 1:}\\
Die folgende Tabelle\footnote{%
In \hyperlink{AppArith3a}{Anhang E zu diesem Kapitel} finden Sie den Quelltext
zur Bestimmung der Tabelle mit Sage.
} zeigt die Werte $a^i{\rm ~mod~}11$ für die Exponenten $i = 1, 2, \cdots, 10$ und für 
die Basen $a = 1, 2, \cdots, 10$ sowie den sich für jedes a daraus ergebenden Wert $ord_{11}(a)$:

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\hypertarget{RSAKonkret}{}%
\hypertarget{Chapter_ElementaryNT_12}{}%ist wahrscheinlich redundant !
\section{Das RSA-Verfahren mit konkreten Zahlen\footnotemark}
\footnotetext{%
    \index{Sage}%
    \index{Minh Van Nguyen}%
    Weiteres Material: Minh Van Nguyen: \glqq Number Theory and the RSA Public Key
    Cryptosystem. Introductory tutorial on using Sage to study elementary number theory
    and public key cryptography'', 2009.
    Didaktisch sehr klarer Artikel zu einigen Grundlagen der Zahlentheorie und zur
    Benutzung von Sage.
    \href{http://nguyenminh2.googlepages.com/sage_numtheory-rsa.pdf}
       {\texttt{http://nguyenminh2.googlepages.com/sage\_numtheory-rsa.pdf}}.
    \vskip +3 pt
}
\label{rsakonkret}\index{RSA}
Nachdem oben die \hyperlink{RSA}{Funktionsweise des RSA-Verfahrens} beschrieben
wurde, sollen diese Schritte hier mit konkreten, aber kleinen Zahlen
durchgeführt werden.

@@

\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: m = 11
sage: for a in xrange(1, m):
....:     print [power_mod(a, i, m) for i in xrange(1, m)]
....:
[1, 1, 1, 1, 1, 1, 1, 1, 1, 1]
[2, 4, 8, 5, 10, 9, 7, 3, 6, 1]
[3, 9, 5, 4, 1, 3, 9, 5, 4, 1]
[4, 5, 9, 3, 1, 4, 5, 9, 3, 1]
[5, 3, 4, 9, 1, 5, 3, 4, 9, 1]
[6, 3, 7, 9, 10, 5, 8, 4, 2, 1]
[7, 5, 2, 3, 10, 4, 6, 9, 8, 1]
[8, 9, 6, 4, 10, 3, 2, 5, 7, 1]
[9, 4, 3, 5, 1, 9, 4, 3, 5, 1]
[10, 1, 10, 1, 10, 1, 10, 1, 10, 1]
\end{Verbatim}
\hypertarget{AppArith3a}{}
\subsection*{Multiplikative Ordnung und Primitivwurzel}
\index{Primitivwurzel}

@@

@@

\vskip +10 pt \hypertarget{AppArith4b}
{{\bf Beispiel auf Seite}} \pageref{SrcArith4b}. \\
Die Faktorisierung von $J(256.027) = 255.016 = 2^3 * 127 * 251$ bestimmt Mathematica mit 
{\tt FactorInteger[255016]= \{\{2,3\}, \{127,1\}, \{251,1\}\}}. \\
In Sage lautet die Syntax: \\
{\tt factor(255016)}.
TODO
\vskip +10 pt \hypertarget{AppArith4b}{%
\noindent {\bf Example on page~\pageref{SrcArith4b}.}} \\
The factorisation of $J(256027) = 255016 = 2^3 * 127 * 251$ can be
calculated using Sage as follows:

@@

\vskip +10 pt \hypertarget{AppArith4c}{}
{{\bf Beispiel auf Seite}} \pageref{SrcArith4c}. \\ 
Mit Mathematica kann die RSA-Verschlüsselung durch den Befehl \\
{\tt PowerMod[\{82, 83, 65, 32, 119, 111, 114, 107, 115, 33\}, 65537, 256027]\}}  \\
berechnet werden.  \\ 
In Sage lautet die Syntax: \\
{\tt vecextract( [Mod(82,256027)\^{}65537, Mod(83,256027)\^{}65537, Mod(65,256027)\^{}65537, \\
                 Mod(32,256027)\^{}65537, Mod(119,256027)\^{}65537, ...] )
}
TODO
\noindent {\bf Example on page~\pageref{SrcArith4c}.}} \\
Sage can do RSA encryption as follows:
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: A = [82, 83, 65, 32, 119, 111, 114, 107, 115, 33]
sage: e = 65537; m = 256027
sage: [power_mod(a, e, m) for a in A]
[212984, 25546, 104529, 31692, 248407, 100412, 54196, 100184, 58179, 227433]
\end{Verbatim}

@@

\vskip +10 pt \hypertarget{AppArith4d}{}
{{\bf Beispiel auf Seite}} \pageref{SrcArith4d}.  \\
Die RSA-Verschlüsselung kann mit Mathematica per \\
{\tt PowerMod[\{21075, 16672, 30575, 29291, 29473\}, 65537, 256027]}  \\
berechnet werden.\\
In Sage lautet die Syntax: \\
{\tt vecextract( [Mod(21075,256027)\^{}65537, Mod(16672,256027)\^{}65537, \\
             Mod(30575,256027)\^{}65537, Mod(29291,256027)\^{}65537, \\
             Mod(29473,256027)\^{}65537], 31) 
}
TODO
\noindent {\bf Example on page~\pageref{SrcArith4d}.}} \\
Sage can do RSA encryption as follows:

@@

\vskip +10 pt \hypertarget{AppArith4e}{}
{{\bf Beispiel auf Seite}} \pageref{SrcArith4e}. \\
Die RSA-Verschlüsselung kann mit Mathematica per \\
{\tt PowerMod[\{82083, 65032, 119111, 114107, 115033\}, 65537, 256027]} \\
berechnet werden. \\
In Sage lautet die Syntax: \\
{\tt vecextract( [Mod(82083,256027)\^{}65537, Mod(65032,256027)\^{}65537, \\
             Mod(119111,256027)\^{}65537, Mod(114107,256027)\^{}65537, \\
             Mod(115033,256027)\^{}65537], 31)
}
TODO
\noindent {\bf Example on page~\pageref{SrcArith4e}.}} \\
RSA encryption using Sage:

>>>>

\hypertarget{Chapter_ElementaryNT}{}
\chapter{Einführung in die elementare Zahlentheorie mit Beispielen}
\label{Chapter_ElementaryNT}
(Bernhard Esslinger, Juli 2001; Updates: Nov. 2001, Juni 2002, Mai 2003,
 Mai 2005, März 2006, Juni 2007) \\

@@

{\bf Beispiel 1:}\\
Die folgende Tabelle\footnote{%
In \hyperlink{AppArith3a1}{Anhang E zu diesem Kapitel} finden Sie den Quelltext
zur Bestimmung der Tabelle mit Sage.
} zeigt die Werte $a^i{\rm ~mod~}11$ für die Exponenten $i = 1, 2, \cdots, 10$ und für 
die Basen $a = 1, 2, \cdots, 10$ sowie den sich für jedes a daraus ergebenden Wert $ord_{11}(a)$:

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\hypertarget{RSAKonkret}{}%
\hypertarget{Chapter_ElementaryNT_12}{}%ist wahrscheinlich redundant !
\section[Das RSA-Verfahren mit konkreten Zahlen]
	{Das RSA-Verfahren mit konkreten Zahlen\footnotemark}
\footnotetext{%
    \index{Sage}%
    \index{Minh Van Nguyen}%
    Weiteres Material: Minh Van Nguyen: \glqq Number Theory and the RSA Public Key
    Cryptosystem. Introductory tutorial on using Sage to study elementary number theory
    and public key cryptography'', 2009.
    Didaktisch sehr klarer Artikel zu einigen Grundlagen der Zahlentheorie und zur
    Benutzung von Sage.
    \href{http://nguyenminh2.googlepages.com/sage_numtheory-rsa.pdf}
       {\texttt{http://nguyenminh2.googlepages.com/sage\_numtheory-rsa.pdf}}.
    \vskip +3 pt
}
\label{rsakonkret}\index{RSA}
Nachdem oben die \hyperlink{RSA}{Funktionsweise des RSA-Verfahrens} beschrieben
wurde, sollen diese Schritte hier mit konkreten, aber kleinen Zahlen
durchgeführt werden.

@@

\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: m = 11
sage: for a in xrange(1, m):
....:     print [power_mod(a, i, m) for i in xrange(1, m)]
....:
[1, 1, 1, 1, 1, 1, 1, 1, 1, 1]
[2, 4, 8, 5, 10, 9, 7, 3, 6, 1]
[3, 9, 5, 4, 1, 3, 9, 5, 4, 1]
[4, 5, 9, 3, 1, 4, 5, 9, 3, 1]
[5, 3, 4, 9, 1, 5, 3, 4, 9, 1]
[6, 3, 7, 9, 10, 5, 8, 4, 2, 1]
[7, 5, 2, 3, 10, 4, 6, 9, 8, 1]
[8, 9, 6, 4, 10, 3, 2, 5, 7, 1]
[9, 4, 3, 5, 1, 9, 4, 3, 5, 1]
[10, 1, 10, 1, 10, 1, 10, 1, 10, 1]
\end{Verbatim}
\hypertarget{AppArith3a1}{}
\subsection*{Multiplikative Ordnung}

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\hypertarget{AppArith3a2}{}
\subsection*{Primitivwurzeln}
\index{Primitivwurzel}
TODO
Computing a primitive root in Sage is very straightforward. If
\verb!n! is an integer, the command \verb!primitive_root(n)! computes
a primitive root of the multiplicative group
$(\mathbf{Z} / n \mathbf{Z})^{\ast}$, if one exists. Where $n$ is
prime, then this is the same as calculating a primitive root of
$\mathbf{Z} / n \mathbf{Z}$. Here, we calculate some primitive roots
of a few integers.

\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: primitive_root(4)
3
sage: primitive_root(22)
13
sage: for p in primes(1, 50):
....:     print p, primitive_root(p)
....:     
2 1
3 2
5 2
7 3
11 2
13 2
17 3
19 2
23 5
29 2
31 3
37 2
41 6
43 3
47 5
\end{Verbatim}

If $p$ is prime, then $\mathbf{Z} / p \mathbf{Z}$ has at least one
primitive root. Sometimes we want to compute all the primitive roots
of $\mathbf{Z} / p \mathbf{Z}$, not just any primitive root of
$\mathbf{Z} / p \mathbf{Z}$. The following function can do this.

\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: def enum_primroots(M):
....:     r"""
....:     Return all the primitive roots of M.
....:     """
....:     g = primitive_root(M)
....:     targetOrder = euler_phi(M)
....:     L = []
....:     # Stepping through all odd integers from 1 up to M, not including
....:     # M. So this loop only considers values of i where 1 <= i < M.
....:     for i in xrange(1, M, 2):
....:         testGen = Mod(g^i, M)
....:         if testGen.multiplicative_order() == targetOrder:
....:             L.append(testGen.lift())
....:     # removing duplicates
....:     return Set(L)
\end{Verbatim}

\noindent For example, here is a list of all primitive roots of the prime 541.

\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: enum_primroots(541)
{2, 517, 10, 523, 13, 14, 527, 528, 18, 531, 24, 539, 30, 37, 40, 51,
54, 55, 59, 62, 65, 67, 68, 72, 73, 77, 83, 86, 87, 91, 94, 96, 98,
99, 107, 113, 114, 116, 117, 126, 127, 128, 131, 132, 138, 150, 152,
153, 156, 158, 163, 176, 181, 183, 184, 195, 197, 199, 206, 208,
210, 213, 218, 220, 223, 224, 244, 248, 250, 257, 258, 259, 260,
261, 263, 267, 269, 270, 271, 272, 274, 278, 280, 281, 282, 283,
284, 291, 293, 297, 317, 318, 321, 323, 328, 331, 333, 335, 342,
344, 346, 357, 358, 360, 365, 378, 383, 385, 388, 389, 391, 403,
409, 410, 413, 414, 415, 424, 425, 427, 428, 434, 442, 443, 445,
447, 450, 454, 455, 458, 464, 468, 469, 473, 474, 476, 479, 482,
486, 487, 490, 501, 504, 511}
\end{Verbatim}

\noindent With a little bit of programming, we can find out how long
it takes to compute all the primitive roots of some number of primes.

\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: def primroots_iter(start, end):
....:     r"""
....:     Compute all primitive roots of all primes between start and end,
....:     inclusive. This uses a primes iterator.
....:     """
....:     for p in primes(start, end+1):
....:         enum_primroots(p)
....:         
sage: %time primroots_iter(1, 1500)
CPU times: user 2.04 s, sys: 0.17 s, total: 2.21 s
Wall time: 2.21 s
\end{Verbatim}

\noindent And we can count how many primes for which to enumerate all
their primitive roots.

\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: def primroots_primetest(start, end):
....:     r"""
....:     Compute all primitive roots of all primes between start and end,
....:     inclusive. This actually perform primality tests, so it allows us
....:     to count the number of primes from start to end, inclusive.
....:     """
....:     nPrimes = 0
....:     for n in xrange(start, end+1):
....:         if is_prime(n):
....:             nPrimes += 1
....:             enum_primroots(n)
....:     print "Enumerate all primitive roots of %s primes." % nPrimes
....:     
sage: %time primroots_primetest(1, 1500)
Enumerate all primitive roots of 239 primes.
CPU times: user 2.35 s, sys: 0.03 s, total: 2.38 s
Wall time: 2.39 s
\end{Verbatim}

Using our custom-defined function \verb!enum_primroots!, we can find
all primitive roots of a prime integer $p$. From this list of primitive
roots, we can determine the smallest and largest primitive root for
$\mathbf{Z} / p \mathbf{Z}$, as well as count the number of primitive
roots of $\mathbf{Z} / p \mathbf{Z}$.
Figure~\ref{fig:primitive_roots_all} graphs the number of primitive
roots for each prime between 1 and 100000. The $x$-axis represents
primes between 1 and 100000, while the $y$-axis counts the number of
primitive roots for each prime within that interval.

\begin{figure}[!htbp]
\centering
\includegraphics{figures/primitive-roots-all}
\caption{The number of primitive roots of all primes between 1 and 100,000.}
\label{fig:primitive_roots_all}
\end{figure}

Figure~\ref{fig:primitive_roots_smallest} graphs the smallest
primitive roots of all primes between 1 and 100000. The $x$-axis
represents primes between 1 and 100000. The $y$-axis represents the
smallest primitive root of each prime within that
interval. Figure~\ref{fig:primitive_roots_largest} shows a
corresponding graph for the largest primitive root of each prime
within the above interval.

\begin{figure}[!htbp]
\centering
\includegraphics{figures/primitive-roots-smallest}
\caption{The smallest primitive roots of all primes between 1 and 100,000.}
\label{fig:primitive_roots_smallest}
\end{figure}

\begin{figure}[!htbp]
\centering
\includegraphics{figures/primitive-roots-largest}
\caption{The largest primitive roots of all primes between 1 and 100,000.}
\label{fig:primitive_roots_largest}
\end{figure}

@@

\vskip +10 pt \hypertarget{AppArith4b}{%
\noindent {\bf Beispiel auf Seite \pageref{SrcArith4b}.}} \\
Die Faktorisierung von $J(256.027) = 255.016 = 2^3 * 127 * 251$ bestimmt Mathematica mit 
{\tt FactorInteger[255016]= \{\{2,3\}, \{127,1\}, \{251,1\}\}}. \\
In Sage lautet die Syntax: \\
{\tt factor(255016)}.
TODO
\vskip +10 pt \hypertarget{AppArith4b}{%
\noindent {\bf Example on page~\pageref{SrcArith4b}.}} \\
The factorisation of $J(256027) = 255016 = 2^3 * 127 * 251$ can be
calculated using Sage as follows:

@@

\vskip +10 pt \hypertarget{AppArith4c}{%
\noindent {\bf Beispiel auf Seite \pageref{SrcArith4c}.}} \\ 
Mit Mathematica kann die RSA-Verschlüsselung durch den Befehl \\
{\tt PowerMod[\{82, 83, 65, 32, 119, 111, 114, 107, 115, 33\}, 65537, 256027]\}}  \\
berechnet werden.  \\ 
In Sage lautet die Syntax: \\
{\tt vecextract( [Mod(82,256027)\^{}65537, Mod(83,256027)\^{}65537, Mod(65,256027)\^{}65537, \\
                 Mod(32,256027)\^{}65537, Mod(119,256027)\^{}65537, ...] )
}
TODO
\noindent {\bf Example on page~\pageref{SrcArith4c}.}} \\
Sage can do RSA encryption as follows:
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: A = [82, 83, 65, 32, 119, 111, 114, 107, 115, 33]
sage: e = 65537; m = 256027
sage: [power_mod(a, e, m) for a in A]
[212984, 25546, 104529, 31692, 248407, 100412, 54196, 100184, 58179, 227433]
\end{Verbatim}

@@

\vskip +10 pt \hypertarget{AppArith4d}{%
\noindent {\bf Beispiel auf Seite \pageref{SrcArith4d}.}} \\
Die RSA-Verschlüsselung kann mit Mathematica per \\
{\tt PowerMod[\{21075, 16672, 30575, 29291, 29473\}, 65537, 256027]}  \\
berechnet werden.\\
In Sage lautet die Syntax: \\
{\tt vecextract( [Mod(21075,256027)\^{}65537, Mod(16672,256027)\^{}65537, \\
             Mod(30575,256027)\^{}65537, Mod(29291,256027)\^{}65537, \\
             Mod(29473,256027)\^{}65537], 31) 
}
TODO
\noindent {\bf Example on page~\pageref{SrcArith4d}.} \\
Sage can do RSA encryption as follows:

@@

\vskip +10 pt \hypertarget{AppArith4e}{%
\noindent {\bf Beispiel auf Seite \pageref{SrcArith4e}.}} \\
Die RSA-Verschlüsselung kann mit Mathematica per \\
{\tt PowerMod[\{82083, 65032, 119111, 114107, 115033\}, 65537, 256027]} \\
berechnet werden. \\
In Sage lautet die Syntax: \\
{\tt vecextract( [Mod(82083,256027)\^{}65537, Mod(65032,256027)\^{}65537, \\
             Mod(119111,256027)\^{}65537, Mod(114107,256027)\^{}65537, \\
             Mod(115033,256027)\^{}65537], 31)
}
TODO
\noindent {\bf Example on page~\pageref{SrcArith4e}.} \\
RSA encryption using Sage:
